Refinements for open automata

### 2023-04-04 2021-Refinement/paper/refinementarticleV2.tex
add Figures

\TODO{Debut Exemple-1}
\begin{example} As a running example, we consider a classical producer-consumer pair  with FIFO buffer. The overall structure shown in Fig. \ref{Fig:Architect}
reflects the different processes involved in the communication and all the actions performed.

\begin{figure}[h]
 \centering
   \includegraphics[width=.70\textwidth]{Figures/Architecture.pdf}
   \caption{Open Automaton. It exposes an unspecified counter in the "cnt" hole
 to count external tick actions. The system sets a counter when a query is enabled and updates  when  the time is over.\label{Fig:Architect}} \TODO{refaire l exemple pour l instant il a pas vraiment de sens}
\end{figure}

The open automaton  modelling the behaviour of the \texttt{prod-cons}  system is depicted in Fig. \ref{Fig:SpecOA}. The \texttt{prod-cons} can be fed using an operation \texttt{push} and discharged  using an operation \texttt{pop}. The automata has two holes: $\texttt{P}$ and  $\texttt{Q}$  depicting  the behaviour of the producer and consumer resp. When the \texttt{prod-cons} agrees with $\texttt{P}$ on a data admission, it   stores the data in a local variable and  exposes to the environment an observable action \texttt{push}. At the opposite side, when it agrees with $\texttt{Q}$ on a data retrieval, it exposes  a \texttt{pop} action. It can also agree with $\texttt{Q}$ on admission of data but this one is rather printed not stored (\texttt{pop(m)}). 
Besides, producing and  consuming, it can agree on the computation action of both processes but this action is exposed as unobservable actions $\tau$.

\medskip

a data base example inspired from \cite{Gorrieri:2001}. The open automaton \texttt{database}  modelling the behaviour of data base is depicted in Fig. \ref{Fig:SpecOA}. The \texttt{database} can be queried using an operation \texttt{qry} and updated using an operation \texttt{upd}. The associated automaton has a single hole:  a counter ($\texttt{cnt}$)  depicting  the behaviour of the timer. When the \texttt{database}  performs a query,  it sets the new time limit of the counter component and the exposed action to the environment is an observable  action \texttt{qry}.
The \texttt{database} updates and  switches when it agrees with the counter component the time is over, and the exposed action is \texttt{upd}.
The open automaton \texttt{counter}    depicting  the behaviour of the timer is illustrated in Fig. \ref{Fig:counter}. 
\end{example}
\begin{figure}[h]
 \centering
   \includegraphics[width=.95\textwidth]{Figures/FIFOpen2.pdf}
   \caption{Open Automaton. It exposes an unspecified counter in the "cnt" hole
 to count external tick actions. The system sets a counter when a query is enabled and updates  when  the time is over.\label{Fig:SpecOA}} \TODO{refaire l exemple pour l instant il a pas vraiment de sens}
\end{figure}

@@

OA are partially specified automata, that partiality comes  from the holes.
A hole can be seen as a port in which we can plug an OA.
The plugging operation is called composition.
The composition of OA was already implicitly defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never  formalised on OA\footnote{The definition of composition below is a direct translation of what happens with pNets composition \cite{henrio:01299562} without  introducing pNets.}

@@

\begin{figure}[h]
 \centering
   \includegraphics[width=.35\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.45\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) Open Automaton of \texttt{counter} process.  Its role is to get a target, then count ticks until that target is reached, then emits an action with the elapsed time and restart. (Right) Open Automaton of the composition \texttt{database}[\texttt{counter/cnt}].\label{Fig:counter}}
\end{figure}

\begin{example}
In Fig. \ref{Fig:counter} (right) we present the composition \texttt{database}[\texttt{counter/cnt}] of the two open automata \texttt{database} and \texttt{counter}. Each state of the composite consists of a state of \texttt{database} together with a state of \texttt{counter}. The composed automaton  takes over  the same steps as the   \texttt{database} automaton but it also includes new steps,  indicating the change of states for the setting of timer.  Note how restricting the composition to a synchronisation of actions (of the encompassing automaton and  of the hole) corresponds to imposing of a change in states. This example also shows how value passing works: the time limit of the counter is set by the parameter \texttt{n} carried by  the \texttt{set} action.
\end{example}

@@

\begin{definition}[Classical Bisimulation]
  A bisimulation is a relation $\mathcal{R}$ such that $s~\mathcal{R}~t$:\\
  \begin{minipage}[c]{.6\textwidth}
    \[
      \forall l~s',~ {s}\xrightarrow{l}{s'}
      \implies
      \exists t'.~ s' ~\mathcal{R}~ t'
      \land {t}\xrightarrow{l}{t'}
    \]
    \begin{center}
      and conversely
    \end{center}
    \vspace{-1ex}
    \[
      \forall l~t',~ {t}\xrightarrow{l}{t'}
      \implies
      \exists s'.~ s' ~\mathcal{R}~ t'
      \land {s}\xrightarrow{l}{s'}
    \]
    \vspace{.5ex}
  \end{minipage}
  i.e. \qquad
  \begin{minipage}[c]{.35\textwidth}
    \begin{tikzpicture}
      \node (s) at (-1,1) {$s$};
      \node (R) at (0,1) {$~\mathcal{R}$};
      \node (t) at (1,1) {$t$};
      \node (sp) at (-1,0) {$s^\prime$}
      edge [<-] node[auto] {$l$} (s);
      \node (Rp) at (0,0) {$\mathcal{R}$};
      \node (tp) at (1,0) {$t^\prime$}
      edge [<-] node[auto,swap] {$l$} (t);
    \end{tikzpicture}
  \end{minipage}
$s$ and $t$ are bisimilar, written $s\tilde t$ iff there is a bisimulation relation $R$ sucht that  $s ~\mathcal{R}~ t$
\end{definition}

@@

\begin{example} 
To illustrate the refinement simulation of open automata, consider a variation on the \texttt{database}.  We suppose that on more concrete level, in addition to querying and updating the data base,  a backup copy of it may also be made. The behaviour of the concrete database is given by 
an open automaton  depicted in Fig. \ref{Fig:RefineOA}. 
\end{example}
\begin{figure}[!tb]
 \centering
   \includegraphics[width=.9\textwidth]{Figures/databaseRefine.pdf}
   \caption{Another variant of the \texttt{database} process.  Two other actions,  in addition to  \texttt{qry} and \texttt{upd},  \texttt{back} representing a request to make a backup copy and    \texttt{cpy} are used to reflect the actual operation of copying. \label{Fig:RefineOA}}
\end{figure}

@@

\begin{figure}[h]
 \centering
   \includegraphics[width=.6\textwidth]{Figures/composeRefined.pdf}
   \caption{Open automata resulting from the composition of the  refined  \texttt{database} and the \texttt{counter} \label{Fig:ComposeRefine}}
\end{figure}

@@

\section{Example}
\TODO{A completer...}

\begin{figure}[!tb]
 \centering
   \includegraphics[width=.65\textwidth]{Figures/Architecture.pdf}
   \caption{\texttt{Archtecture du systeme}. \label{Fig:Archi}}
\end{figure}

\begin{figure}[!tb]
 \centering
   \includegraphics[width=.90\textwidth]{Figures/FIFOpen2.pdf}
   \caption{\texttt{FIFO} process.  Two other actions,  in addition to  \texttt{p-a} and \texttt{p-b},  \texttt{in} representing a     \texttt{out} are used to reflect the actual operation of copying. \label{Fig:FIFOA}}
\end{figure}

%\section{Proof of Transitivity for Refinement}\label{sec:proof-transitivity}

\begin{figure}[!tb]
 \centering
   \includegraphics[width=1.25\textwidth]{Figures/FIFORing.pdf}
   \caption{\texttt{FIFO} process.  \label{Fig:RefineOA}}
\end{figure}

\begin{figure}[h]
 \centering
   \includegraphics[width=.40\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.55\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) Open Automaton of \texttt{P} process.   (Right) Open Automaton of  \texttt{Q} process.\label{Fig:processes}}
\end{figure}

>>>>

\TODO{Debut Exemple-1}
\newpage
\begin{example} As a running example, we consider a classical producer-consumer pair  with FIFO buffer,  referred to as \texttt{prod-cons}. The  structure shown in Fig. \ref{Fig:Architect} reflects the different processes involved in the system and all the actions performed.

\begin{figure}[h]
 \centering
   \includegraphics[width=.70\textwidth]{Figures/Architecture.pdf}
   \caption{The structure of the example \label{Fig:Architect}} 
\end{figure}

The open automaton  modelling the behaviour of the \texttt{prod-cons}  system is depicted in Fig. \ref{Fig:SpecOA}. The \texttt{prod-cons} can be fed using an operation \texttt{push} and discharged  using an operation \texttt{pop}. The automata has two holes: $\texttt{P}$ and  $\texttt{Q}$  depicting  the behaviour of the producer and consumer resp. When the \texttt{prod-cons} agrees with $\texttt{P}$ on a data admission, it   stores the data in a local variable and  exposes to the environment an observable action \texttt{push}. At the opposite side, when it agrees with $\texttt{Q}$ on a data retrieval, it exposes  a \texttt{pop} action. The system can also agree with $\texttt{Q}$ on the admission of data but this one is rather printed not stored.  Besides, producing and  consuming, it can agree on the computation action of both processes but this action is exposed as unobservable actions $\tau$.

\begin{figure}[h]
 \centering
   \includegraphics[width=.95\textwidth]{Figures/FIFOpen2.pdf}
   \caption{Open Automaton of the \texttt{prod-cons} system. It exposes two unspecified processes in holes
   $\texttt{P}$ and  $\texttt{Q}$. \label{Fig:SpecOA}} \TODO{refaire l exemple pour l instant il a pas vraiment de sens}
\end{figure}

\end{example}

@@

OA are partially specified automata, that partiality comes  from the holes.
A hole can be seen as a port in which we can plug an OA.
The plugging operation is called composition.
The composition of OA was already implicitly defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never  formalised on OA.\footnote{The definition of composition below is a direct translation of what happens with pNets composition \cite{henrio:01299562} without  introducing pNets.}

@@

Two processes that are composable with the open automaton in  Fig. \ref{Fig:SpecOA} are those given in  Fig. \ref{Fig:procandcons}. 
\begin{figure}[h]
 \centering
   \includegraphics[width=.35\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.45\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) Open Automaton of the \texttt{producer}.  It  transmits data calculated, sending one item at a time to the FIFO. (Right) The \texttt{consumer} gets one item at a time from the FIFO and sends it after a prior processing. \label{Fig:procandcons}}
\end{figure}
~~\\
\smallskip
\RAB{Pbme: L'automate de la fig \ref{Fig:ComposeOA} n'est pas le resultat de la composition du prod et cons avec la sepc, i.e l'automate de la fifure \ref{Fig:SpecOA} mais avec celui du raffine (qui vient apres) i.e fig \ref{Fig:RefineOA}. Solutions: 1) deplacer la composition 2) refaire la figure l'automate sera un peu plus gros? }
\begin{figure}[h]
 \centering
   \includegraphics[width=.95\textwidth]{Figures/compos-FIFO-Pro.pdf}
   \caption{Open Automaton of the \texttt{prod-cons}[\texttt{P}/\texttt{producer}]. 
   $\texttt{P}$ and  $\texttt{Q}$. \label{Fig:ComposeOA}} 
\end{figure}
\end{example}

@@

\begin{definition}[Classical Bisimulation]
  A bisimulation is a relation $\mathcal{R}$ such that $s~\mathcal{R}~t$:\\
  \begin{minipage}[c]{.6\textwidth}
    \[
      \forall l~s',~ {s}\xrightarrow{l}{s'}
      \implies
      \exists t'.~ s' ~\mathcal{R}~ t'
      \land {t}\xrightarrow{l}{t'}
    \]
    \begin{center}
      and conversely
    \end{center}
    \vspace{-1ex}
    \[
      \forall l~t',~ {t}\xrightarrow{l}{t'}
      \implies
      \exists s'.~ s' ~\mathcal{R}~ t'
      \land {s}\xrightarrow{l}{s'}
    \]
    \vspace{.5ex}
  \end{minipage}
  i.e. \qquad
  \begin{minipage}[c]{.35\textwidth}
    \begin{tikzpicture}
      \node (s) at (-1,1) {$s$};
      \node (R) at (0,1) {$~\mathcal{R}$};
      \node (t) at (1,1) {$t$};
      \node (sp) at (-1,0) {$s^\prime$}
      edge [<-] node[auto] {$l$} (s);
      \node (Rp) at (0,0) {$\mathcal{R}$};
      \node (tp) at (1,0) {$t^\prime$}
      edge [<-] node[auto,swap] {$l$} (t);
    \end{tikzpicture}
  \end{minipage}
$s$ and $t$ are bisimilar, written $s\sim t$ iff there is a bisimulation relation $R$ such that  $s ~\mathcal{R}~ t$
\end{definition}

@@

\begin{example} 
To illustrate the refinement simulation of open automata, consider a variation on the \texttt{prod-cons}.  We suppose that on more concrete level,
FIFO circular buffer
  The behaviour of the concrete database is given by 
an open automaton  depicted in Fig. \ref{Fig:RefineOA}. 
\end{example}
\begin{figure}[!tb]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/FIFORing.pdf}
   \caption{Open Automation of the \texttt{prod-cons} system communicating through a FIFO circular buffer.  \label{Fig:RefineOA}}
\end{figure}

@@

%\begin{figure}[h]
% \centering
%   \includegraphics[width=.6\textwidth]{Figures/composeRefined.pdf}
%   \caption{Open automata resulting from the composition of the  refined  \texttt{database} and the \texttt{counter} \label{Fig:ComposeRefine}}
%\end{figure}

@@
